A Hexagon Tile Map Algorithm for Displaying Spatial Data

### 2020-10-21 kobakian-cook.tex
Editing to end of doc

\hypertarget{user-choices}{%
\paragraph{User choices}\label{user-choices}}

@@

A user can tweak the parameters of the hexagon map using the additional
arguments to the \texttt{create\_hexmap} function, these options may
affect the speed of the creation.

\hypertarget{creating-a-hexagon-tile-map}{%
\paragraph{Creating a hexagon tile
map}\label{creating-a-hexagon-tile-map}}

The following code creates the hexagon tile map for all the Statistical
Areas at Level 2 in Tasmania. These steps can be executed by the main
function, \texttt{create\_hexmap}, or can be implemented separately for
more flexibility.

If a user would like to perform each of the steps of the algorithm
themselves, the necessary inputs will change for each function.

Users may choose to only use the \texttt{allocate} function when they
wish to use a set of centroids, rather than \citep{sf} polygons.

\hypertarget{polygon-set}{%
\paragraph{Polygon set}\label{polygon-set}}

The polygon set of Statistical Areas at Level 2 (SA2) \citep{abs2011} of
Tasmania in 2011 is provided with the \texttt{sugarbag} package as
\texttt{tas\_sa2}. A single column of the data set is used to identify
the unique areas. In this case, the unique SA2 names for each SA2 have
been used.

The longitude and latitude centre of the capital cities of Australia are
used as focal points to allocate each geographic area around the closest
capital city. Hobart will be the common focal point, as this example
uses only the areas in the state of Tasmania.

@@

\hypertarget{step-1-derive-the-set-of-centroid-points}{%
\paragraph{Step 1: Derive the set of centroid
points}\label{step-1-derive-the-set-of-centroid-points}}

A set of centroids may be used directly. The set of polygons should be
provided as an \texttt{sf} object, this is a data frame containing a
\texttt{geometry} column. The \texttt{read\_shape} function can assist
in creating this object for use in \texttt{R}.

@@

\hypertarget{step-2-create-the-hexagon-grid-points}{%
\paragraph{Step 2: Create the hexagon grid
points}\label{step-2-create-the-hexagon-grid-points}}

A grid is created to ensure tessellation between the hexagons that
represent the geographic units on a hexagon tile map.

The grid of possible hexagon locations is made using the
\texttt{create\_grid} function. It uses the centroids, the hexagon size
and the buffer distance.

\hypertarget{creating-a-tessellated-grid}{%
\subparagraph{Creating a tessellated
grid}\label{creating-a-tessellated-grid}}

A set of longitude columns, and latitude rows are created to define the
locations of the hexagons. The distance between each row and column is
the size specified by \texttt{hex\_size}. Equally spaced columns are
created from the minimum longitude minus the buffer distance, up to the
maximum longitude plus the buffer distance. Similarly, the rows are
created from the latitude values and the buffer distance. A unique
hexagon location is created from all intersections of the longitude
columns and latitude rows. Figure \ref{fig:filterprocess} shows the
original grid on the left, to allow for tessellating hexagons, every
second latitude row on the grid is shifted right, by half of the hexagon
size. The grid for tessellation is shown on the right.

\hypertarget{rolling-windows}{%
\subparagraph{Rolling windows}\label{rolling-windows}}

Not all of the grid points will be used, especially if islands result in
a large grid space. To filter the grid for appropriate hexagon locations
for allocation, the \texttt{create\_buffer} function is used by
\texttt{create\_grid}. It finds the grid points needed to best capture
the set of centroids on a hexagon tile map.

The closest latitude row and longitude column are found for each
centroid location. Then rows and columns of centroids are divided into
20 groups. The amount of rows in each latitude group and the amount of
columns in each longitude group are used as the width of rolling
windows. The rolling windows can be seen on the bottom and right of the
grid shown in Figure \ref{fig:filterprocess}. This will tailor the
available grid points to those most likely to be used. It also helps
reduce the amount of time taken, as it decreases the amount of points
considered for each centroid allocation.

The first rolling window function finds the minimum and maximum centroid
values for the sliding window groups of longitude columns and the groups
of latitude rows.

The second rolling window function finds the average of the rolling
minimum and maximum centroid values, for the longitude columns and
latitude rows.

\hypertarget{filtering-the-grid}{%
\subparagraph{Filtering the grid}\label{filtering-the-grid}}

The grid points are kept only if they fall between the rolling average
of the minimum and maximum centroid values after accounting for the
buffer distance, for each row and column of the grid. The sparsely
populated South-West region of National Park has much fewer points
available compared to the South-East region containing the city of
Hobart.

\hypertarget{centroid-to-focal-point-distance}{%
\subparagraph{Centroid to focal point
distance}\label{centroid-to-focal-point-distance}}

The distance between each centroid in the set, and each of the focal
points provided is calculated. The name of the closest focal point, and
the distance and angle from focal point to polygon centroid is joined to
polygon data set. To minimise time taken for this step only one option
is provided, Tasmania's capital city Hobart. The order for allocation is
determined by the distance between the polygon centroid and it's closest
focal point. The points are arranged from the centroid closest to the
focal point(s), to the furthest.

\hypertarget{step-3-allocate-each-centroid-to-a-hexagon-grid-point}{%
\paragraph{Step 3: Allocate each centroid to a hexagon grid
point}\label{step-3-allocate-each-centroid-to-a-hexagon-grid-point}}

Allocation of all centroids takes place using the set of polygon
centroids and the hexagon map grid. Centroid allocation begins with the
closest centroid to a focal point. This will preserve spatial
relationships with the focal point, as the inner city areas are
allocated first, they will be placed closest to the capital, and the
areas that are further will then be accommodated. The possible hexagon
grid points reduces by one after each allocation, then only those that
have not yet been allocated are considered.

@@

\hypertarget{filter-the-grid-for-unassigned-hexagon-points}{%
\subparagraph{Filter the grid for unassigned hexagon
points}\label{filter-the-grid-for-unassigned-hexagon-points}}

@@

\hypertarget{filter-the-grid-points-for-those-closest-to-the-centroid}{%
\subparagraph{Filter the grid points for those closest to the
centroid}\label{filter-the-grid-points-for-those-closest-to-the-centroid}}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=1.2\linewidth]{kobakian-cook_files/figure-latex/filterprocess-1} \caption[Filter for grid points within a square, then circular, distance for those closest to the centroid]{Filter for grid points within a square, then circular, distance for those closest to the centroid.}\label{fig:filterprocess}
\end{figure}
\end{Schunk}

@@

\hypertarget{hexagon-tile-map-of-australia}{%
\subsubsection{Hexagon tile map of
Australia}\label{hexagon-tile-map-of-australia}}

This algorithm has been applied to the complete set of all Australian
Statistical Areas at Level 2. This display highlights the density of
Australian capital cities, as it draws attention to the many communities
in Sydney, Melbourne and Hobart. This display also highlights the
disparity in the burden of thyroid cancer for females in the communities
of these three cities. There are several collections of red hexagons in
Hobart that draw attention and they represent the communities with much
higher rates of diagnosis than the Australian average. The communities
south of Hobart show a gradient of colour as the values progress from
higher than average, to much lower than average for the communities
closer to the Sydney CBD. This pattern progresses into the communities
with lower than average rates in Melbourne and Tasmania.

Compared to the choropleth map display, the low rates in the rural
Australian communities are no longer dominating the display. The much
higher than average rates in Sydney draw more attention in the hexagon
tile map display unlike the red and orange areas in the city of Hobart
that draw attention in both displays.

@@

The \texttt{gganimate} \citep{gganimate} package can be used to make an
animation. It requires connecting the polygons for each area in two
displays, which can be done using the \texttt{sf\_id} variable, such as
the SA2 name. The animation connecting these two displays will highlight
the rapid growth of the inner-city areas that emphasises the density of
the communities. The hexagons that move the furthest will move rapidly
in the animation. The rapid decreases of the large rural areas also show
how much greater the landmass of Statistical Areas can be.

@@

It is possible to use alternative maps to communicate spatial
distributions.While a choropleth map display is the current practice
spatial visualisation of geographical data. Current methods do not
always work for Australia due to the large geographic space between the
densely populated capital cities. The administrative boundaries may also
distract from the statistics communicated using colour.

Alternative maps highlight the value of the statistics across the
geographic units. Alternative mapping methods allow increased
understanding of the spatial distribution of a variable across the
population, by fairly representing each administrative area. This
acknowledges that the amount of residents can be different but
recognises that each population area is equally important. The solution
to this visualisation problem has equally sized areas, with
neighbourhood boundary connections. This map algorithm is implemented in
the \texttt{sugarbag} \citep{sugarbag} package written for \texttt{R}.
The \texttt{sugarbag} package creates tessellated hexagon tile maps. The
Australian application preserves the spatial relationships, emphasising
capital cities. The hexagon tile map is a visualisation solution that
highlights spatial distributions.

These hexagons equally represent each area. However, the tessellation
does not allow the size of the hexagons to represent another variable,
similar to the choropleth maps. The algorithm is heavily dependent on
the focal points used, as this determines the order of allocation. It
works on the assumption that viewers can use directional relationships
to identify their neighbourhoods but this can be aided by the animation.
With careful consideration of the choropleth map, the small geographic
inner city areas may have been noticed by viewers, but the hexagon tile
map display emphasises them. The communities in northern Tasmania and
the Northern territory do not draw attention because of their size as in
the choropleth, but their colour is still noticeably below average when
contrasted with the hexagons further south.

The Dorling cartogram was the most influential in the design of the
hexagon tile map. The use of a hexagon provides a simple shape that
shifts the focus of the visualisation from the shape of SA2s to the
distribution effects of a disease on communities. not impeded by the
irregular and unusual shapes created by a contiguous cartogram and the
immense white space created on the non-contiguous cartogram. Hexagons
can emphasise the relationship between neighbouring communities by
allowing their sides to touch. The tessellation crated by adjoining
hexagons for densely populated areas is helpful as it shows that most
communities are not islands onto themselves, but are intrinsically
connected to those around them. As seen in the Dorling cartogram in
figure \ref{fig:tas_displays}c), the tessellation that allows boundary
connections was not imposed for rural communities, where centroid of the
geographic unit was located far from others, a feature that allows
distant but dense populations of the coastal towns of Tasmania to be
recognised.

The choropleth map that uses the geographic shapes of areas can be seen
in figure \ref{fig:choro}. Figure \ref{fig:hexmap} displays the
Australian areas with equally sized hexagons. This display allows the
small and densely populated inner city areas to be easily contrasted
with their neighbouring areas.

Future work will include refining the algorithm. It would be possible to
take a logarithmic function rather than a direct angle to help choose a
closer hexagon to the original centroid location, before increasing the
width of the angle used to filter the hexagons.

This algorithm has only been tested using single countries, and does not
consider definite borders of countries. While the buffer allows
extension beyond the furthest centroids, there is no mechanism to
protect the borders and ensure centroids are placed within the
geographic borders of a country.

This algorithm is an effective start to creating hexagon tile maps for
many geographic units.

\hypertarget{installation}{%
\paragraph{Installation}\label{installation}}

@@

@@

Load the library into your R session with:

>>>>

\hypertarget{user-choices}{%
\subsubsection{User choices}\label{user-choices}}

@@

A user can tweak the parameters of the hexagon map using the additional
arguments to the \texttt{create\_hexmap} function, these options may
affect the speed of the creation. The hexagon size may need adjustments
depending on the density of the population, users can provide an
appropriate hexagon size and re-run the algorithm. The buffer distance
may need to be increased if the coastal cities need to extend beyond the
geographic land mass.

The following code creates the hexagon tile map for all the Statistical
Areas at Level 2 in Tasmania. These steps can be executed by the main
function, \texttt{create\_hexmap}, or can be run separately for more
flexibility.

If a user would like to perform each of the steps of the algorithm
themselves, the necessary inputs will change for each function. For
example, the \texttt{allocate} function can be used directly when a user
wishes to use a set of centroids, rather than \citep{sf} polygons.

The polygon set of Statistical Areas at Level 2 (SA2) \citep{abs2011} of
Tasmania in 2011 is accessed from the \texttt{absmapsdata} package. A
single column of the data set is used to identify the unique areas. In
this case, the unique SA2 names for each SA2 have been used.

The longitude and latitude centre of the capital cities of Australia are
used as focal points to allocate each geographic area around the closest
capital city. Hobart will be the focal point, as this example uses only
the areas in the state of Tasmania.

@@

\hypertarget{step-1-derive-the-set-of-centroid-points}{%
\subsubsection{Step 1: Derive the set of centroid
points}\label{step-1-derive-the-set-of-centroid-points}}

The set of polygons should be provided as an \texttt{sf} object, this is
a data frame containing a \texttt{geometry} column. The
\texttt{read\_shape} function can assist in creating this object for use
in \texttt{R}.

@@

\hypertarget{step-2-create-the-hexagon-grid-points}{%
\subsubsection{Step 2: Create the hexagon grid
points}\label{step-2-create-the-hexagon-grid-points}}

A grid is created to ensure tessellation between the hexagons that
represent the geographic units on a hexagon tile map. The grid of
possible hexagon locations is made using the \texttt{create\_grid}
function. It uses the centroids, the hexagon size and the buffer
distance.

\hypertarget{creating-a-tessellated-grid}{%
\paragraph{Creating a tessellated
grid}\label{creating-a-tessellated-grid}}

A set of longitude columns, and latitude rows are created to define the
locations of the hexagons. The distance between each row and column is
the size specified by \texttt{hex\_size}. Equally spaced columns are
created from the minimum longitude minus the buffer distance, up to the
maximum longitude plus the buffer distance. Similarly, the rows are
created from the latitude values and the buffer distance. A unique
hexagon location is created from all intersections of the longitude
columns and latitude rows. Figure \ref{fig:filterprocess}a) allows for
tessellating hexagons after every second latitude row on the grid is
shifted right, by half of the hexagon size.

\hypertarget{rolling-windows}{%
\paragraph{Rolling windows}\label{rolling-windows}}

Not all of the grid points will be used, especially if islands result in
a large grid space. To filter the grid for appropriate hexagon locations
for allocation, the \texttt{create\_grid} function will call the
\texttt{create\_buffer} function. It finds the grid points needed to
best capture the set of centroids on a hexagon tile map. It also helps
reduce the amount of time taken, as it decreases the amount of points
considered for each centroid allocation.

The closest latitude row and longitude column are found for each
centroid location. Then rows and columns of centroids are divided into
20 groups. The amount of rows in each latitude group and the amount of
columns in each longitude group are used as the width of rolling
windows.

The first rolling window step finds the minimum and maximum centroid
values within each of the sliding window groups of longitude columns,
and the groups of latitude rows.

The second rolling window step finds the average of the rolling minimum
and maximum centroid values, for the longitude columns and latitude
rows.

\hypertarget{filtering-the-grid}{%
\paragraph{Filtering the grid}\label{filtering-the-grid}}

The hexagon grid points are kept only if they fall within the rolling
average of the minimum and maximum centroid values after accounting for
the buffer distance, for each row and column of the grid. The sparsely
populated South-West region of National Park has very few points
available compared to the South-East region near the city of Hobart.

\hypertarget{centroid-to-focal-point-distance}{%
\paragraph{Centroid to focal point
distance}\label{centroid-to-focal-point-distance}}

The distance is calculated between each centroid in the set, and each of
the focal points provided. The distance and angle from the closest focal
point to polygon centroid is used in the allocation process as the order
for allocation is determined by the distance between the polygon
centroid and it's closest focal point.

\hypertarget{step-3-allocate-each-centroid-to-a-hexagon-grid-point}{%
\subsubsection{Step 3: Allocate each centroid to a hexagon grid
point}\label{step-3-allocate-each-centroid-to-a-hexagon-grid-point}}

Allocation of all centroids takes place using the set of polygon
centroids and the hexagon map grid. The points are arranged from the
centroid closest to the focal point(s), to the furthest. Centroid
allocation begins with the closest centroid to a focal point. This will
preserve spatial relationships with the focal point, as the inner city
areas are allocated first, they will be placed closest to the capital,
and the areas that are further will then be accommodated. The set of
possible hexagon grid points reduces by one after each centroid is
allocated, then only the grid points that have not yet been selected are
considered.

@@

\hypertarget{filter-the-grid-for-unassigned-hexagon-points}{%
\paragraph{Filter the grid for unassigned hexagon
points}\label{filter-the-grid-for-unassigned-hexagon-points}}

@@

\hypertarget{filter-the-grid-points-for-those-closest-to-the-centroid}{%
\paragraph{Filter the grid points for those closest to the
centroid}\label{filter-the-grid-points-for-those-closest-to-the-centroid}}

@@

\begin{Schunk}
\begin{figure}
\includegraphics[width=0.95\linewidth]{kobakian-cook_files/figure-latex/filterprocess-1} \caption[Filter for grid points within a square, then circular, distance for those closest to the centroid]{Filter for grid points within a square, then circular, distance for those closest to the centroid.}\label{fig:filterprocess}
\end{figure}
\end{Schunk}

@@

When the algorithm is applied to the complete set of all Australian
Statistical Areas at Level 2 it results in the hexagon tile map display
shown in Figure \ref{fig:hexmap}. This display highlights the density of
Australian capital cities, as it draws attention to the many communities
in Sydney, Melbourne and Hobart. This display also highlights the
disparity in the burden of thyroid cancer for females in the communities
of these cities. There are several collections of red hexagons in Sydney
that represent the communities with much higher rates of diagnosis than
the Australian average. Brisbane also experiences higher than average
rates of diagnosis, but has more orange than red, which represents much
higher rates. The females in the cities of Adelaide and Perth show much
lower rates of diagnosis.

Compared to the choropleth map display, the low rates in the rural
Australian communities are no longer dominating the display. While the
representative hexagons are still visible against the black background,
the much lower than average rates in the Northern Territory are less
noticeable.

@@

Creating an animation connecting these two displays can allow users t
grasp the meaning of the connected hexagons. It will also highlight the
density of the communities using the rapid expansion of the inner-city
areas, these hexagons will move the furthest and will move rapidly in
the animation from their geographic locations. The rapid decrease in
size of the large rural areas can show how large the landmass of
Statistical Areas can be, and how few people occupy rural areas. The
\texttt{gganimate} \citep{gganimate} package can be used to make an
animation. It connects the polygons for each area in the two displays
using the \texttt{sf\_id} variable, the names of the statistical areas,
and creates intermediate displays as they geographic areas shrink or
expand to the chosen hexagon size.

@@

A choropleth map display is the current practice spatial visualisation
for geographical data. It is possible to enhance the communication of
spatial distributions using alternative map displays. Current methods
for alternative maps do not always work for Australia due to the large
rural geographic space between the densely populated capital cities. The
administrative boundaries may also distract from the statistics
communicated using colour due to the small size of some areas.

Alternative maps highlight the value of the statistics across the
geographic units. These methods allow increased understanding of the
spatial distribution of a variable across the population, by fairly
representing each administrative area. This acknowledges that the amount
of residents can be different but recognises that each population area
is equally important. The solution to this visualisation problem can be
using equally sized areas, with neighbourhood boundary connections. The
\texttt{sugarbag} package for \texttt{R} creates tessellated hexagon
tile maps by automating an algorithm to design these displays. The
Australian application preserves the spatial relationships, emphasising
capital cities.

These hexagons equally represent each area. However, similar to the
choropleth map display the tessellation does not allow the size of the
hexagons to represent another variable. The algorithm is heavily
dependent on the focal points used, as this determines the order of
allocation. It works on the assumption that viewers can identify their
neighbourhoods using the relative directional relationships but this can
be aided by the use of animation. With careful consideration of the
choropleth map, the small geographic inner city areas may have been
noticed by viewers, but the hexagon tile map display emphasises them.
The communities in northern Tasmania and the Northern territory do not
draw attention because of their size as in the choropleth, but their
colour is still noticeably below average when contrasted with the
hexagons in New South Wales.

The Dorling cartogram was the most influential alternative map display
for the design of the hexagon tile map. The use of a hexagon shifts the
focus of the visualisation from the shape of SA2s to the distribution
effects of a disease on communities. It is not impeded by the irregular
and unusual shapes created by a contiguous cartogram and the immense
white space created on the non-contiguous cartogram. Hexagons can
emphasise the relationship between neighbouring communities by allowing
their sides to touch. The tessellation created by adjoining hexagons for
densely populated areas is helpful as it shows that most communities are
not islands onto themselves, but are intrinsically connected to those
around them. As seen in the Dorling cartogram in figure
\ref{fig:tas_displays}c), the tessellation that allows boundary
connections was not imposed for rural communities, where centroid of the
geographic unit was located far from others, a feature that allows
distant but dense populations of the coastal towns of southern Tasmania
to be recognised.

This algorithm has only been tested using single countries, and does not
consider definite borders of countries. While the buffer allows
extension beyond the furthest centroids, there is no mechanism to
protect the borders and ensure centroids are placed within the
geographic borders of a country. It would be also possible to take a
logarithmic function in the filter steps for selection of a hexagon grid
point. A direct angle is currently used, but this may narrowly miss
appropriate hexagons, a logarithmic function may help to choose a closer
hexagon to the original centroid location, before increasing the width
of the angle used to filter the hexagons. This algorithm is an effective
start to creating hexagon tile maps for many geographic units.

\hypertarget{installation}{%
\subsubsection{Installation}\label{installation}}

@@

\begin{Schunk}
\begin{Sinput}
install.packages("sugarbag")
\end{Sinput}
\end{Schunk}

@@

\begin{Schunk}
\begin{Sinput}
devtools::install_github("srkobakian","sugarbag")
\end{Sinput}
\end{Schunk}
